---
title: "TDDE07: Lab 1 report"
author: "Sebastian Callh (sebca553), Jacob Lundberg (jaclu010)"
date: "10 april 2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(fig.align = "center")
posterior.col  = "orange"
likelihood.col = "red"
prior.col      = "green"
```

## 1. Bernoulli ... again

### a)
The mean of a Bernoulli distribution is given by $p = \frac{s}{n} = \frac{14}{20} \approx 0.7$ which is very close to the mean of the simulated posterior distribution plotted below. 

```{r}
# a) Show that posterior converges to 14/20 approx. 0.7
alpha0 <- 2
beta0  <- 2 
nDraws <- 10000
n      <- 20
s      <- 14
f      <- n - s
post   <- rbeta(nDraws, alpha0 + s, beta0 + f)
hist(post, 
     breaks = 100,
     main   = "Histogram of posterior draws",
     xlab   = expression(theta))
```

The posterior mean and the true mean are close to each other and increasing the number of draws will let them come arbitrarily close.

```{r}
sprintf("Mean of posterior draws %f", mean(post))
```

### b)
The posterior probability and the true probability for $P(\theta < 0.4)$ lie very close to each other. As with the mean value, increasing the number of draws will let the posterior probability approach the true probability.

```{r}
# b) Simulation to compute posterior probability P(theta < 0.4 | y)
thetas <- rbeta(nDraws, alpha0 +s , beta0 + f)
sprintf("Simulated probability of theta < 0.4: %f", length(thetas[thetas<0.4]) / length(thetas))  # Simulated value
sprintf("Exact probability of theta < 0.4: %f", pbeta(0.4, alpha0 + s, beta0 + f))                # Exact value
```

### c)
The log-odds posterior distribution can be seen in the plot below.

```{r}
# c) Log odds posterior distribution
phi <- sapply(thetas, function(theta) { log(theta / (1 - theta)) })
phi.dens <- density(phi)
plot(phi.dens,
     main = "Log-odds posterior distribution", 
     ylab = "Density",
     xlab = expression(phi),
     col  = posterior.col)
```

\pagebreak

## 2. Log-normal distribution and the Gini coefficient. 

### a)
Simulating from the posterior and plotting the approximated density together with the theoretical density reveals that the approximation is quite good, which can be seen in the plot below.

```{r}
incomes <- c(14, 25, 45, 25, 30, 33, 19, 50, 34, 67)
mu      <- 3.5
n       <- length(incomes)
tau2    <- sum((log(incomes) - mu)^2) / n 
nDraws  <- 10000
dinvgamma <- function(x, a, b) {
  (b^a)/gamma(a) * x^(-a-1) * exp(-b/x)
}

dinvchisq2 <- function(x, n, tau2) {
  a <- n / 2
  b <- n*tau2 / 2
  dinvgamma(x, a, b)
}

# a) Simulate posterior draws
X           <- rchisq(nDraws, n)
sigma2      <- n*tau2/X
sigma2.dens <- density(sigma2)

plot(sigma2.dens, col = "blue", 
     main = "Posterior density of sigma^2",
     xlab = expression(sigma^2),
     ylim = c(0,5),
     xlim = c(0,2))

# Theoretical sigmas
delta  <- 0.01
grid   <- seq(0.01, 2.5, delta)
sigmas <- dinvchisq2(grid, n, tau2)
lines(grid, sigmas/(sum(sigmas)*delta), type = 'l', col = "orange")
legend("topright", 
       c("Simulated distribution", "Theoretical distribution"), 
       col = c("blue", "orange"),
       lty = 1,
       bty='n', 
       cex=.75)
```

### b)
Using the samples from a) and computing the posterior distribution of the Gini coefficient G results in the plot below.

```{r}
# b) Gini coefficient posterior
G      <- 2*pnorm(sqrt(sigma2/2)) - 1
#G.hist <- hist(G, breaks = 100, freq = FALSE, xlim = 0:1)

# c) G Credible interval
G.slice <- sort(G)[(0.025*nDraws):(0.975*nDraws)]
cred.lower.bound <- min(G.slice)
cred.upper.bound <- max(G.slice)
G.dens <- density(G)
plot(G.dens,
     main = "Posterior distribution of G",
     xlab = "G",
     col  = posterior.col)
```

### c)
Computing the 95% credible interval results in 
```{r}
sprintf("Credible interval lower bound: %f", cred.lower.bound)
sprintf("Credible interval upper bound: %f", cred.upper.bound)
```
and computing the Highest Posterior Density interval results in

```{r}
# c) G Highest Posterior Density (HPD)
i      <- order(G.dens$y, decreasing = TRUE)
csum   <- cumsum(G.dens$y[i])
m      <- length(csum[csum < sum(G.dens$y)*0.95])
H      <- G.dens$x[i][1:m]
hpd.lower.bound <- min(H)
hpd.upper.bound <- max(H)
sprintf("HPD lower bound: %f", hpd.lower.bound)
sprintf("HPD upper bound: %f", hpd.upper.bound)
```

\pagebreak

## 3. Bayesian inference for the concentration parameter in the Von Mises distribution
Plotting the posterior distribution of $\kappa$ together with the prior and likelihood results in the plot below.

```{r}

Von.Mises <- function (y, mu, k){
  exp(k*cos(y - mu)) / (2*pi*besselI(k, 0))
}

Y          <- c(-2.44, 2.14, 2.54, 1.83, 2.02, 2.33, -2.79, 2.23, 2.07, 2.02)
mu         <- 2.39
lambda     <- 1
delta      <- 0.05
kappas     <- seq(0, 10, delta)
likelihood <- sapply(kappas, function (k) { prod(Von.Mises(Y, mu, k)) })
prior      <- dexp(kappas, lambda)
posterior  <- likelihood * prior

plot(kappas, posterior/(sum(posterior)*delta),
     main = "Posterior distribution of kappa",
     xlab = expression(kappa),
     ylab = "Density",
     col  = posterior.col, 
     type = 'l')
lines(kappas, likelihood/(sum(likelihood)*delta), col ="red")
lines(kappas, prior/(sum(prior)*delta), col ="green")
legend("topright", 
       c("Posterior", "Likelihood", "Prior"), 
       lty=1, 
       col=c(posterior.col, likelihood.col, prior.col), 
       bty='n', 
       cex=.75)

```

Which has it's mode at
```{r}
sprintf("Posterior mode: %.2f", kappas[which.max(posterior)])
```

\pagebreak

```{r, echo=TRUE, eval=FALSE}
# 1
# a) Show that posterior converges to 14/20 approx. 0.7
alpha0 <- 2
beta0  <- 2 
nDraws <- 10000
n      <- 20
s      <- 14
f      <- n - s
post   <- rbeta(nDraws, alpha0 + s, beta0 + f)
hist(post, 
     breaks = 100,
     main   = "Histogram of posterior draws",
     xlab   = expression(theta))

sprintf("Mean of posterior draws %f", mean(post))

# b) Simulation to compute posterior probability P(theta < 0.4 | y)
thetas <- rbeta(nDraws, alpha0 +s , beta0 + f)
sprintf("Simulated probability of theta < 0.4: %f", length(thetas[thetas<0.4]) / length(thetas))  # Simulated value
sprintf("Exact probability of theta < 0.4: %f", pbeta(0.4, alpha0 + s, beta0 + f))                # Exact value

# c) Log odds posterior distribution
phi <- sapply(thetas, function(theta) { log(theta / (1 - theta)) })
phi.dens <- density(phi)
plot(phi.dens,
     main = "Log-odds posterior distribution", 
     ylab = "Density",
     xlab = expression(phi),
     col  = posterior.col)

# 2
incomes <- c(14, 25, 45, 25, 30, 33, 19, 50, 34, 67)
mu      <- 3.5
n       <- length(incomes)
tau2    <- sum((log(incomes) - mu)^2) / n 
nDraws  <- 10000
dinvgamma <- function(x, a, b) {
  (b^a)/gamma(a) * x^(-a-1) * exp(-b/x)
}

dinvchisq2 <- function(x, n, tau2) {
  a <- n / 2
  b <- n*tau2 / 2
  dinvgamma(x, a, b)
}

# a) Simulate posterior draws
X           <- rchisq(nDraws, n)
sigma2      <- n*tau2/X
sigma2.dens <- density(sigma2)

plot(sigma2.dens, col = "blue", 
     main = "Posterior density of sigma^2",
     xlab = expression(sigma^2),
     ylim = c(0,5),
     xlim = c(0,2))

# Theoretical sigmas
delta  <- 0.01
grid   <- seq(0.01, 2.5, delta)
sigmas <- dinvchisq2(grid, n, tau2)
lines(grid, sigmas/(sum(sigmas)*delta), type = 'l', col = "orange")
legend("topright", 
       c("Simulated distribution", "Theoretical distribution"), 
       col = c("blue", "orange"),
       lty = 1,
       bty='n', 
       cex=.75)

# b) Gini coefficient posterior
G      <- 2*pnorm(sqrt(sigma2/2)) - 1
#G.hist <- hist(G, breaks = 100, freq = FALSE, xlim = 0:1)

# c) G Credible interval
G.slice <- sort(G)[(0.025*nDraws):(0.975*nDraws)]
cred.lower.bound <- min(G.slice)
cred.upper.bound <- max(G.slice)
G.dens <- density(G)
plot(G.dens,
     main = "Posterior distribution of G",
     xlab = "G",
     col  = posterior.col)

sprintf("Credible interval lower bound: %f", cred.lower.bound)
sprintf("Credible interval upper bound: %f", cred.upper.bound)

# c) G Highest Posterior Density (HPD)
i      <- order(G.dens$y, decreasing = TRUE)
csum   <- cumsum(G.dens$y[i])
m      <- length(csum[csum < sum(G.dens$y)*0.95])
H      <- G.dens$x[i][1:m]
hpd.lower.bound <- min(H)
hpd.upper.bound <- max(H)
sprintf("HPD lower bound: %f", hpd.lower.bound)
sprintf("HPD upper bound: %f", hpd.upper.bound)


# 3

Von.Mises <- function (y, mu, k){
  exp(k*cos(y - mu)) / (2*pi*besselI(k, 0))
}

Y          <- c(-2.44, 2.14, 2.54, 1.83, 2.02, 2.33, -2.79, 2.23, 2.07, 2.02)
mu         <- 2.39
lambda     <- 1
delta      <- 0.05
kappas     <- seq(0, 10, delta)
likelihood <- sapply(kappas, function (k) { prod(Von.Mises(Y, mu, k)) })
prior      <- dexp(kappas, lambda)
posterior  <- likelihood * prior

plot(kappas, posterior/(sum(posterior)*delta),
     main = "Posterior distribution of kappa",
     xlab = expression(kappa),
     ylab = "Density",
     col  = posterior.col, 
     type = 'l')
lines(kappas, likelihood/(sum(likelihood)*delta), col ="red")
lines(kappas, prior/(sum(prior)*delta), col ="green")
legend("topright", 
       c("Posterior", "Likelihood", "Prior"), 
       lty=1, 
       col=c(posterior.col, likelihood.col, prior.col), 
       bty='n', 
       cex=.75)

sprintf("Posterior mode: %.2f", kappas[which.max(posterior)])
```

