---
title: "TDDE07 - Lab 4"
author: "Sebastian Callh, Jacob Lundberg"
date: "16 may 2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
source('tdde07_lab4_sebca553_jaclu010_code.R')
```

## Poisson regression - the MCMC way

#a)

Using `glm` to estimate $\beta$ resulted in the coefficients in the table below.

\begin{table}[h!]
  \begin{center}
    \caption{Point estimate of $\beta$.}
    \begin{tabular}{c|c|c|c|c|c|c|c|c}
    (Intercept) & PowerSeller &  VerifyID   &   Sealed  &   Minblem &    MajBlem    & LargNeg &    LogBook & MinBidShare \\ 
    \hline
     1.072      & -0.021      & -0.395      & 0.444     & -0.052    & -0.221        & 0.071    &    -0.121 &   -1.894 \\
    \end{tabular}
  \end{center}
\end{table}

It can be seen that the coefficients  `VerifyID`, `Sealed`, `MajBlem`, `MinBidShare` are important, with `MinBidShare` being most important. 

#b)

Using `optim` to compute the posterior mode (and mean) of the coefficients for the normal approximation resulted in the coefficients in the table below. 

\begin{table}[h!]
  \begin{center}
    \caption{Point estimate of $\beta$.}
    \begin{tabular}{c|c|c|c|c|c|c|c|c}
    (Intercept) & PowerSeller &  VerifyID   &   Sealed  &   Minblem &    MajBlem    & LargNeg &    LogBook & MinBidShare \\ 
    \hline
    1.070 & -0.021 & -0.393 &  0.444 & -0.052 & -0.221 &  0.071 & -0.120&  -1.892 \\
    \end{tabular}
  \end{center}
\end{table}

#c)

Using Metropolis random walk to simulate from the posterior $\beta$ distribution resulted in the posterior mean coefficients which can be seen below.

\begin{table}[h!]
  \begin{center}
    \caption{Point estimate of $\beta$.}
    \begin{tabular}{c|c|c|c|c|c|c|c|c}
    (Intercept) & PowerSeller &  VerifyID   &   Sealed  &   Minblem &    MajBlem    & LargNeg &    LogBook & MinBidShare \\ 
    \hline
    1.068   & -0.017 & -0.394 &  0.445 & -0.051 & -0.224 &  0.076 & -0.121 &  -1.890 \\
    \end{tabular}
  \end{center}
\end{table}

```{r}
plot.posterior.beta.samples()
```

Observing the plots above it can be seen that all chains have converged to the static distribution.

```{r}
plot.posterior.phis()
```

Mapping onto $\phi = \exp(\beta)$ and plotting the densities gives the plots that can be seen above.

```{r}
plot.graphical.comparison()
```

Finally comparing the differenet estimated $\beta$ shows that they all did a good job.

#d)

Using the MCMC draws from c) to simulate from the predictive distribution of the provided data point resulted in the distribution that can be seen below. The expected value of no bidders was $0.36$.

```{r}
plot.posterior.predictive.density()
```

\pagebreak

```{r, code=readLines('tdde07_lab4_sebca553_jaclu010_code.R'), eval=FALSE,echo=TRUE}

```

