---
title: "TDDE07 - Lab 3"
author: "Sebastian Callh, Jacob Lundberg"
date: "4 maj 2018"
output:
  pdf_document:
    fig_height: 4
    fig_width: 5
    fig_caption: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(fig.align = "center")
source('lab3_no_stan.R')
```

## 1. Normal model, mixture of normal model with semi-conjugate prior

### a) - Normal model

#### (ii) - Analysis
Looking at the trajectories for $\mu$ and $\sigma$ it can be seen that the values are not very auto correlated, which implies good draws. They also oscillate around a stable value.
```{r}
plot.gibbs.trajectory()
```

### b) Mixture normal model
After running the `NormalMixtureGibbs.R` for 30 iterations the mixture model had converged to the distribution seen in the plot below.
\centerline{\includegraphics[width=0.9\textwidth]{mixture-of-normals.eps}}

### c) Graphical comparison
\centerline{\includegraphics[width=0.9\textwidth]{graphical-compasison.eps}}

## 2. Time-series models in Stan

### a)
Plotting realizations of the AR(1)-process shows that a high $\phi$ gives a smoothing effect and a small (into the negatives) gives an oscillating effect. When $\phi$ is close to zero all contributions of $x_{t-1}$ is supressed and only white noise remains.

```{r}
plot.ar.process()
```

### b)

### i)
For each of the processes, $\sigma$ 
4000 draws.

\begin{table}[h!]
  \begin{center}
    \caption{Estimated parameters of AR(1) process X ($\phi = 0.3$)}
    \begin{tabular}{c|c|c|c}
      & $\mu$ & $\sigma$ & $\phi$ \\
      \hline
      Mean  & 7.25 & 1.38 & 0.28 \\
      2.5\%  & 5.83 & 1.24 & 0.14 \\
      97.5\% & 8.62 & 1.53 & 0.42 \\
      Eff. samples & 1108 & 1450 & 1098 \\
      True value & 10 & 1.41 & 0.3 \\
    \end{tabular}
  \end{center}
\end{table}

\centerline{\includegraphics[width=0.7\textwidth]{estimated-parameters-of-from-x.eps}}

\begin{table}[h!]
  \begin{center}
    \caption{Estimated parameters of AR(1) process Y ($\phi = 0.95$)}
    \begin{tabular}{c|c|c|c}
      & $\mu$ & $\sigma$ & $\phi$ \\
      \hline
      Mean   & 0.35  & 1.34 & 0.96 \\
      2.5\%  & -0.10 & 1.22 & 0.92 \\
      97.5\% & 0.81  & 1.48 & 1.00 \\
      Eff. samples & 1726 & 2166 & 1729 \\
      True value & 10 & 1.41 & 0.95 \\
    \end{tabular}
  \end{center}
\end{table}

\centerline{\includegraphics[width=0.7\textwidth]{estimated-parameters-of-from-y.eps}}

### ii)

\centerline{\includegraphics[width=0.9\textwidth]{joint-posterior-density-ar-y.eps}}
\centerline{\includegraphics[width=0.9\textwidth]{joint-posterior-density-ar-x.eps}}

### c)






\centerline{\includegraphics[width=0.9\textwidth]{posterior-credible-interval.eps}}

\centerline{\includegraphics[width=0.9\textwidth]{posterior-without-sigma-prior.eps}}


\centerline{\includegraphics[width=0.9\textwidth]{posterior-with-sigma-0.02-prior.eps}}

\centerline{\includegraphics[width=0.9\textwidth]{posterior-with-sigma-prior.eps}}


### d)

\pagebreak

## Appendix
```{r, code=readLines('tdde07_lab3_sebca553_jaclu010_code.R'), echo=TRUE, eval=FALSE}

```

### Stan file

```{r, echo=TRUE, eval=FALSE}
data {
  int<lower=0> N;
  vector[N] x;
}

parameters {
  real mu;
  real<lower=0> sigma;
  real phi;
}

model {
  x[2:N] ~ normal(mu + phi * x[1:(N - 1)], sigma);
}
```

