---
title: "TDDE07 - Lab 3"
author: "Sebastian Callh, Jacob Lundberg"
date: "4 maj 2018"
output:
  pdf_document:
    fig_height: 4
    fig_caption: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
source('lab3_no_stan.R')
```

## 1. Normal model, mixture of normal model with semi-conjugate prior

### a) - Normal model

#### (i) - Gibbs sampler
```{r}
plot.normal.approximation()
```


#### (ii) - Analysis

### b) Mixture normal model


### c) Graphical comparison
```{r}
plot.graphical.comparison()
```

\pagebreak

## 2. Time-series models in Stan

### a)
```{r}
plot.ar.process()
```

### b)


### c)

\pagebreak

!["Posterior without sigma prior"](posterior-without-sigma-prior.eps)

\pagebreak

!["Posterior with sigma 0.02 prior"](posterior-with-sigma-0.02-prior.eps)

\pagebreak

!["Posterior with sigma prior"](posterior-with-sigma-prior.eps)

\pagebreak

### d)

\pagebreak

## Appendix
```{r, code=readLines('tdde07_lab3_sebca553_jaclu010_code.R'), echo=TRUE, eval=FALSE}

```

